Normalization by Evaluation for Sized Dependent Types

### 2017-02-20 nbe/icfp17.tex
[ ICFP 17 ] Fixed some calls to Rne.
[ ICFP 17 ] Some cases of the fundamental theorem.

We extend the syntax by de Bruijn levels $\lev k$
to be used as generic values (unknowns) and type annotations
$\up T t$ and $\down T t$ for lazy realisations of the
reflection and reification operations of NbE.
\[
\begin{array}{lllrl@{\qquad}l}
%  \Exp  & \ni & t & ::= & \dots \mid \lev k \mid \up T t \mid \down T t \\
  \Ne   & \ni & n & ::= & \dots \mid \lev k \\
  \tUp  & \ni & N & ::= & \up T n \\
  \Whnf & \ni & w & ::= & \dots \mid N \mid \down T t \\
%  \Whnf & \ni & w,d,f,g,A,B,F,G & ::= & \dots \mid  \up T n \mid \down T t \\
\end{array}
\]

@@

Judgements.
\[
\begin{array}{l@{\qquad}l}
   \R k t  v     & \mbox{} \\
   \RTy k T  V  & \mbox{} \\
   \RNat k t  v & \mbox{} \\
   \RNe k t  m  & \mbox{} \\
\end{array}
\]
\fbox{$\R k t v$}
\begin{gather*}
  \ru{t \evalsto \down{U} T \qquad
      U \evalsto s \\
      \RTy k T V
    }{\R k t V}
\qquad
  \ru{t \evalsto \down{U} u \qquad
      U \evalsto \Nat a \\
      \RNat k u v
    }{\R k t v}
\qquad
  \ru{t \evalsto \down{U} u \qquad
      U \evalsto \up {T'} {t'} \\
      \RNe k u v
    }{\R k t v}
\\[2ex]
  \ru{t \evalsto \down{U} f \qquad
      U \evalsto \piT A B \qquad
      \R{k+1} {\down{B[\up A \lev k]} (f\,\up A \lev k)} v
    }{\R k t {\lambda v}}
\\[2ex]
  \ru{t \evalsto \down{U} f \qquad
      U \evalsto \piT \Size B \qquad
      \R{k+1} {\down{B[\lev k]} (f\,\lev k)} v
    }{\R k t {\lambda v}}
\qquad
  \ru{t \evalsto \down{U} f \qquad
      U \evalsto \forallT B \qquad
      \R{k+1} {\down{B[\lev k]} (f\,\ann{\lev k})} v
    }{\R k t {\lambda v}}
\end{gather*}

\fbox{$\RNat k t v $}
\begin{gather*}
  \ru{t \evalsto n \qquad
      \RNe k n m
    }{\RNat k t m}
\qquad
  \ru{t \evalsto \zero{a_v} \qquad
      \RSize k {a_v} a
    }{\RNat k t {\zero a}}
\qquad
  \ru{t \evalsto \suc{a_v}u \qquad
      \RSize k {a_v} a \qquad
      \RNat k u v
    }{\RNat k t {\suc a v}}
\end{gather*}

\fbox{$\RSize k {a_v} a$}
\begin{gather*}
  \ru{
    }{\RSize k \infty \infty}
\qquad
  \ru{
    }{\RSize k o o}
\qquad
  \ru{
    }{\RSize k {\lev j + o} {\ind{k-1-j} + o}}
\end{gather*}

\fbox{$\RNe k t m$} and \fbox{$\RElim k e e^v$}
\begin{gather*}
%   \ru{t \evalsto \lev j
%     }{\RNe k t {\ind{k-1-j}}}
% \qquad
  \ru{t \evalsto \lev j\,\vec e \qquad
      \RElim k {e_i} {e^v_i} \mforall i
    }{\RNe k t {\ind{k-1-j}\,\vec e^v}}
\qquad
% \end{gather*}
% \begin{gather*}
  \ru{\R k t v
    }{\RElim k t v}
\qquad
  \ru{\RSize k a b
    }{\RElim k a b}
\qquad
  \ru{\RSize k a b
    }{\RElim k {\ann a} {\ann b}}
\\[2ex]
  \ru{\RTy k T V \qquad
      \R k {t_z} {v_z} \qquad
      \R k {t_s} {v_s}
    }{\RElim k {(\case[\ell] T {t_z} {t_s})} {\case[\ell] V {v_z} {v_s}}}
\qquad
  \ru{\RTy k T V \qquad
      \R k t v
    }{\RElim k {(\fix[\ell] T t)} {\fix[\ell] V v}}
\end{gather*}
\fbox{$\RTy k T V$}
\begin{gather*}
  \ru{T \evalsto \Set[\ell]
    }{\RTy k T {\Set[\ell]}}
\qquad
  \ru{T \evalsto \Nat a \qquad
      \RSize k a b
    }{\RTy k T {\Nat b}}
\qquad
  \ru{T \evalsto \up{U}u \qquad
      \RNe k u m
    }{\RTy k T m}
\\[2ex]
  \ru{T \evalsto \piT A B \qquad
      \RTy k A {V_a} \qquad
      \RTy {k+1} B {V_b}
    }{\RTy k T {\piT {V_a} V_b}}
\qquad
  \ru{T \evalsto \epiT \Size B \qquad
      \RTy {k+1} B {V_b}
    }{\RTy k T {\epiT \Size V_b}}
\end{gather*}

@@

The PERs $\NE$ and $\NF$ charaterize (neutral) normalizing terms.
\[
\begin{array}{l@{\quad}l@{\quad}l@{~}l@{~}l@{~}l@{~}l@{~}l}
  \fbox{$t = t' \in \NE$}
    & \defas &
    \RNe k t m &\mand& \RNe k {t'} {m'} &\mand& m \approx m' &\mforall k \\
  \fbox{$t = t' \in \NF$}
    & \defas &
    \R k t v &\mand& \R k {t'} {v'} &\mand& v \approx v' &\mforall k \\
\end{array}
  % \ru{\RNe k t m \mand \RNe k {t'} {m'} \mand m \approx m' \mforall k
  %   }{t = t' \in \NE}
  % \ru{t \evalsto \up N n \qquad t' \evalsto \up{N'}{n'} \qquad n \approx n'
  %   }{t = t' \in \NE}
\]
All PERs $\A$ we consider are sandwiched $\NE \subseteq \A \subseteq \NF$ between $\NE$ and $\NF$.

@@

By induction on $\ell \in \NN$ we define the PER family
$\_ = \_ \in \Set[\ell]$ of types
together with the extension $\EL[\ell] T$ (for $T = T' \in \Set[\ell]$)
which is PER of values of type $A$.
The rules for \fbox{$T = T' \in \SET[\ell]$} are:
\[
\begin{array}{l@{\qquad}l}
  \ru{T \evalsto N \qquad T' \evalsto N'
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \NE
\\[3ex]
  \ru{T \evalsto \Nat\,a \qquad T' \evalsto \Nat\,a
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \NAT(a)
\\[3ex]
  \rux{T \evalsto \Set[\ell'] \qquad T' \evalsto \Set[\ell']
     }{T = T' \in \SET[\ell]
     }{\ell'{<}\ell}
  & \EL[\ell](T) = \SET[\ell']
\\[3ex]
  \rul{T \evalsto \piT A B \qquad
      T' \evalsto \piT{A'}{B'} \\
      A = A' \in \SET[\ell] \qquad
      B[u] = B'[u'] \in \SET[\ell] \mforall u = u' \in \EL[\ell](A)
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \PI(\EL[\ell](A),\;u \mapsto \EL[\ell](B[u]))
\\[3ex]
  \rul{T \evalsto \piT \Size B \quad
      T' \evalsto \piT{\Size}{B'} \quad
      B[a] = B'[a] \in \SET[\ell] \mforall a \in \SIZE
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \PI(\SIZE,\; a \mapsto \EL[\ell](B[a]))
\\[3ex]
  \rul{T \evalsto \forallT B \quad
      T' \evalsto \forallT{B'} \quad
      B[a] = B'[a] \in \SET[\ell] \mforall a \in \SIZE
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \FORALL(a \mapsto \EL[\ell](B[a]))
\end{array}
\]

\section{Metatheory}
\label{sec:meta}

\begin{lemma}[Typing of constructed data]
\label{lem:invdata} \bla
If\/ $\Gamma \der c : \Nat b$ then $\resurrect\Gamma \der b : \Size$ and $b > 0$.
\end{lemma}

>>>>

We extend the syntax by de Bruijn levels $\lev k$
to be used as generic values (unknowns) and type annotations
$\up T t$ and $\down T t$ for lazy realisations of the
reflection and reification operations of NbE.
\[
\begin{array}{lllrl@{\qquad}l}
%  \Exp  & \ni & t & ::= & \dots \mid \lev k \mid \up T t \mid \down T t \\
  \Ne   & \ni & n & ::= & \dots \mid \lev k \\
  \tUp  & \ni & N & ::= & \up T n \\
  \Whnf & \ni & w & ::= & \dots \mid N \\
  \tDown & \ni & d & ::= & \down T t \\
  \Exp  & \ni & t & ::= & \dots \mid d \\
%  \Whnf & \ni & w,d,f,g,A,B,F,G & ::= & \dots \mid  \up T n \mid \down T t \\
\end{array}
\]

@@

Judgements.
\[
\begin{array}{l@{\qquad}l}
   \R k d  v     & \mbox{} \\
   \RTy k T  V  & \mbox{} \\
   \RNat k t  v & \mbox{} \\
   \RNe k t  m  & \mbox{} \\
\end{array}
\]
\fbox{$\R k d v$}
\begin{gather*}
  \ru{U \evalsto s \qquad
      \RTy k T V
    }{\R k {\down{U} T} V}
\qquad
  \ru{U \evalsto \Nat a \qquad
      \RNat k u v
    }{\R k {\down{U} u} v}
\qquad
  \ru{U \evalsto N \qquad
      u \evalsto \up{T} n \qquad
      \RNe k n m
    }{\R k {\down{U} u} m}
\\[2ex]
  \ru{U \evalsto \piT A B \qquad
      \R{k+1} {\down{B[\up A \lev k]} (f\,\up A \lev k)} v
    }{\R k {\down{U} f} {\lambda v}}
\\[2ex]
  \ru{U \evalsto \piT \Size B \qquad
      \R{k+1} {\down{B[\lev k]} (f\,\lev k)} v
    }{\R k {\down{U} f} {\lambda v}}
\qquad
  \ru{U \evalsto \forallT B \qquad
      \R{k+1} {\down{B[\lev k]} (f\,\ann{\lev k})} v
    }{\R k {\down{U} f} {\lambda v}}
\end{gather*}

\fbox{$\RNat k t v $}
\begin{gather*}
  \ru{t \evalsto \up {\Nat a} n \qquad
      \RNe k n m
    }{\RNat k t m}
\qquad
  \ru{t \evalsto \zero{a_v} \qquad
      \RSize k {a_v} a
    }{\RNat k t {\zero a}}
\qquad
  \ru{t \evalsto \suc{a_v}u \qquad
      \RSize k {a_v} a \qquad
      \RNat k u v
    }{\RNat k t {\suc a v}}
\end{gather*}

\fbox{$\RSize k {a_v} a$}
\begin{gather*}
  \ru{
    }{\RSize k \infty \infty}
\qquad
  \ru{
    }{\RSize k o o}
\qquad
  \ru{
    }{\RSize k {\lev j + o} {\ind{k\dotminus(1+j)} + o}}
\end{gather*}

\fbox{$\RNe k n m$} and \fbox{$\RElim k e e^v$}
\begin{gather*}
%   \ru{t \evalsto \lev j
%     }{\RNe k t {\ind{k-1-j}}}
% \qquad
  \ru{%t \evalsto \lev j\,\vec e \qquad
      \RElim k {e_i} {e^v_i} \mforall i
    }{\RNe k {\lev j\,\vec e} {\ind{k \monus (1+j)}\,\vec e^v}}
\qquad
% \end{gather*}
% \begin{gather*}
  \ru{\R k d v
    }{\RElim k d v}
\qquad
  \ru{\RSize k a b
    }{\RElim k a b}
\qquad
  \ru{\RSize k a b
    }{\RElim k {\ann a} {\ann b}}
\\[2ex]
  \ru{\RTy k T V \qquad
      \R k {d_z} {v_z} \qquad
      \R k {d_s} {v_s}
    }{\RElim k {(\case[\ell] T {d_z} {d_s})} {\case[\ell] V {v_z} {v_s}}}
\qquad
  \ru{\RTy k T V \qquad
      \R k d v
    }{\RElim k {(\fix[\ell] T d)} {\fix[\ell] V v}}
\end{gather*}
\fbox{$\RTy k T V$}
\begin{gather*}
  \ru{T \evalsto \Set[\ell]
    }{\RTy k T {\Set[\ell]}}
\qquad
  \ru{T \evalsto \Nat a \qquad
      \RSize k a b
    }{\RTy k T {\Nat b}}
\qquad
  \ru{T \evalsto \up{U}n \qquad
      \RNe k n m
    }{\RTy k T m}
\\[2ex]
  \ru{T \evalsto \piT A B \qquad
      \RTy k A {V_a} \qquad
      \RTy {k+1} B {V_b}
    }{\RTy k T {\piT {V_a} V_b}}
\qquad
  \ru{T \evalsto \epiT \Size B \qquad
      \RTy {k+1} B {V_b}
    }{\RTy k T {\epiT \Size V_b}}
\end{gather*}

@@

The PERs $\NE$ characterizes neutral reifyable terms.
% \[
% \begin{array}{lll}%{l@{\quad}l@{\quad}l@{~}l@{~}l@{~}l@{~}l@{~}l}
%   \fbox{$\up T n = \up {T'} {n'} \in \NE$}
%     & \defas &
%     T \evalsto N \mand T' \evalsto N' \mand
%     \mforall k \mboth
%     \RNe k n m \mand \RNe k {n'} {m'} \mand m \approx m'
%     .\\
% \end{array}
% \]
The PERs $\NE$ and $\NF$ characterize (neutral) normalizing terms.
\[
\begin{array}{l@{\quad}l@{\quad}l@{~}l@{~}l@{~}l@{~}l@{~}l}
  \fbox{$n = n' \in \NE$}
    & \defas &
    \RNe k n m &\mand& \RNe k {n'} {m'} &\mand& m \approx m' &\mforall k \\
  \fbox{$d = d' \in \NF$}
    & \defas &
    \R k d v &\mand& \R k {d'} {v'} &\mand& v \approx v' &\mforall k \\
\end{array}
  % \ru{\RNe k t m \mand \RNe k {t'} {m'} \mand m \approx m' \mforall k
  %   }{t = t' \in \NE}
  % \ru{t \evalsto \up N n \qquad t' \evalsto \up{N'}{n'} \qquad n \approx n'
  %   }{t = t' \in \NE}
\]
%All PERs $\A$ we consider are sandwiched $\NE \subseteq \A \subseteq \NF$ between $\NE$ and $\NF$.

\begin{lemma}[Closure properties of $\NE$ and $\NF$]
  \bla
  \begin{enumerate}
  \item $\lev k = \lev k \in \NE$.
  \item If $n = n' \in \NE$ and $d = d' \in \NF$ then $n\,d = n'\,d' in \NE$.
%  \item If $d = d' \in \NF$
  \end{enumerate}
\end{lemma}

\[
\fbox{$\NEE$} \defas \{ (\up T n, \up{T'}{n'})
  \msuchthat T \evalsto N \mand T' \evalsto N' \mand n = n' \in \NE \}
\]
\fbox{$\NAT(a)$}
\begin{gather*}
  \ru{t \evalsto \zero a \qquad
      t' \evalsto \zero{a'}
    }{t = t' \in \NAT(b+1)}
\qquad
  \ru{t \evalsto \suc a u \qquad
      t' \evalsto \suc {a'} {u'} \qquad
      u = u' \in \NAT(b)
    }{t = t' \in \NAT(b+1)}
\\[2ex]
  \ru{t \evalsto \up{\Nat a} n \qquad
      t' \evalsto \up{\Nat a'} n' \qquad
      n = n' \in \NE
    }{t = t' \in \NAT(b)}
\end{gather*}

\fbox{$\SIZE$} is a discrete PER of size values:
\begin{gather*}
  \ru{
    }{\infty = \infty \in \SIZE}
\qquad
  \ru{
    }{o = o \in \SIZE}
\qquad
  \ru{
    }{\lev k + o = \lev k + o \in \SIZE}
\end{gather*}

Let $\A$ be a PER and $\F$ a family of PERs over $\A$ such that
$\F(u) = \F(u')$ whenever $u = u' \in \A$.  We define
\[
  \fbox{$\PIAF$} \defas \{(t,t') \mid t\,u = t'\,u' \in \F(u) \mforall u = u' \in \A \}
  .
\]
This works also for $\A = \SIZE$.
For a family $\F$ over $\SIZE$ we also have the irrelevant function space
\[
  \fbox{$\FORALL\,\F$} \defas
  \{(t,t') \mid t \ann a = t' \ann{a'} \in \F(b) \mforall a,a',b \in \SIZE \}
  .
\]

@@

By induction on $\ell \in \NN$ we define the PER family
$\_ = \_ \in \Set[\ell]$ of types
together with the extension $\EL[\ell] T$ (for $T = T' \in \Set[\ell]$)
which is PER of values of type $A$.
The rules for \fbox{$T = T' \in \SET[\ell]$} are:
\[
\begin{array}{l@{\qquad}l}
  \ru{T \evalsto N \qquad T' \evalsto N'
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \NEE
\\[3ex]
  \ru{T \evalsto \Nat\,a \qquad T' \evalsto \Nat\,a
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \NAT(a)
\\[3ex]
  \rux{T \evalsto \Set[\ell'] \qquad T' \evalsto \Set[\ell']
     }{T = T' \in \SET[\ell]
     }{\ell'{<}\ell}
  & \EL[\ell](T) = \SET[\ell']
\\[3ex]
  \rul{T \evalsto \piT A B \qquad
      T' \evalsto \piT{A'}{B'} \\
      A = A' \in \SET[\ell] \qquad
      B[u] = B'[u'] \in \SET[\ell] \mforall u = u' \in \EL[\ell](A)
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \PI(\EL[\ell](A),\;u \mapsto \EL[\ell](B[u]))
\\[3ex]
  \rul{T \evalsto \piT \Size B \quad
      T' \evalsto \piT{\Size}{B'} \quad
      B[a] = B'[a] \in \SET[\ell] \mforall a \in \SIZE
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \PI(\SIZE,\; a \mapsto \EL[\ell](B[a]))
\\[3ex]
  \rul{T \evalsto \forallT B \quad
      T' \evalsto \forallT{B'} \quad
      B[a] = B'[a] \in \SET[\ell] \mforall a \in \SIZE
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \FORALL(a \mapsto \EL[\ell](B[a]))
\end{array}
\]
All PERs involved here are closed under weak head expansion.

The PER model has the usual properties, one of them being that $\EL[\ell](T)$ does not depend $\ell$ nor the derivation that introduced $T = T' \in \SET[\ell]$.  Thus, we simply write $t = t' \in \EL(T)$ or even $t = t' \in T$.

\subsection{Type shapes}

Reflection and reification perform $\eta$-expansion such that we arrive at an $\eta$-long $\beta$-normal form.  To perform the $\eta$-expansion, not the precise type is needed, just the approximate shape, in particular, whether it is a function type (do expand) or a base type (do not expand).
For the logical framework, the shape of a dependent type is just its underlying simple type \cite{harperPfenning:equivalenceLF}. However, in the presence of universes and large eliminations, there is no underlying simple type.  Of course, we can take a type as its own shape, but we want at least that $\Nat a$ and $\Nat b$ have the same shape even for different $a, b$.  Thus, we define a relation $T \shape S$ to express that $S$ is a possible shape of type $T$.

\fbox{$T \shape S$} for $T \in \SET[\ell]$.
\begin{gather*}
  \ru{T \evalsto \Nat a \qquad
      S \evalsto \Nat b
    }{T \shape S}
\qquad
  \ru{T \evalsto N \qquad
      S \evalsto N'
    }{T \shape S}
\qquad
   \ru{T \evalsto \Set[\ell] \qquad
       S \evalsto \Set[\ell]
    }{T \shape S}
\\[2ex]
  \ru{T \evalsto \piT A B \qquad
      S \evalsto \piT{A'}{B'} \qquad
      A \shape A' \qquad
      B[u] \shape B'[u'] \mforall u = u' \in A
    }{T \shape S}
\\[2ex]
  \ru{T \evalsto \piT \Size B \qquad
      S \evalsto \piT{\Size}{B'} \qquad
      B[a] \shape B'[a] \mforall a \in \SIZE
    }{T \shape S}
\\[2ex]
  \ru{T \evalsto \forallT B \qquad
      S \evalsto \forallT{B'} \qquad
      B[a] \shape B'[a'] \mforall a,a' \in \SIZE
    }{T \shape S}
\end{gather*}
% The last rule is the crucial one
\begin{lemma}[Reflexivity for shapes]
  If $T = T' \in \SET[\ell]$ then $T \shape T'$.
\end{lemma}

\begin{theorem}[Reification and reflection]
Let $T \in \SET[\ell]$ and $T \shape S_1$ and $T \shape S_2$.
\begin{enumerate}
\item If $n_1 = n_2 \in \NE$ then $\up{S_1}{n_1} = \up{S_2}{n_2} \in T$.
\item If $t_1 = t_2 \in T$ then $\down{S_1}{t_1} = \down{S_2}{t_2} \in \NF$.
\end{enumerate}
\end{theorem}
\begin{proof}
By induction on $T \in \SET[\ell]$ and cases on $T \shape S_1$ and $T \shape S_2$.
\begin{caselist}

\nextcase $T \evalsto \piT A B$
\begin{gather*}
  \ru{A \in \SET[\ell] \qquad
      B[u] = B[u'] \in \SET[\ell] \mforall u = u' \in A
    }{T \in \SET[\ell]}
\\[2ex]
  \ru{S_1 \evalsto \piT{A_1}{B_1} \qquad
      A \shape A_1 \qquad
      B[u] \shape B_1[u'] \mforall u = u' \in A
    }{T \shape S_1}
\\[2ex]
  \ru{S_2 \evalsto \piT{A_2}{B_2} \qquad
      A \shape A_2 \qquad
      B[u] \shape B_2[u'] \mforall u = u' \in A
    }{T \shape S_2}
\end{gather*}
\begin{enumerate}
\item To show $\up{S_1}{n_1} = \up{S_2}{n_2} \in T$ assume arbitrary $u_1 = u_2 \in A$.
  Let $d_i = \down{A_i}{u_i}$.
  By induction hypothesis, $d_1 = d_2 \in \NF$.
  Thus, $n_1\,d_1 = n_2\,d_2 \in \NE$, and
  by induction hypothesis,  $\up{B_1[u_1]}(n_1\,d_1) = \up{B_2[u_2]}(n_2\,d_2) \in B[u_1]$.
  With $(\up{S_i}{n_i})\,u_i \evalsto \up{B_i[u_i]}(n_i\,d_i)$ we are done by definition of $\EL(T)$.

\item We assume $k \in \NN$ and show $\R k {\down{S_i}{t_i}} \lambda v_i$ for some normal forms
$v_1,v_2$.  Let $u_i = \up{A_i} \lev k$.  Note that $u_1 = u_2 \in A$ by induction hypothesis.
It is sufficient to show
$\R{k+1}{\down{B_i[u_i]}{(t_i\,u_i)}} {v_i}$.
By definition of $\EL(T)$ we have $t_1\,u_1 = t_2\,u_2 \in B[u_1]$, thus, by induction hypothesis,
$\down{B_1[u_1]}(t_1\,u_1) = \down{B_2[u_2]}(t_2\,u_2) \in \NF$, which delivers $v_1$ and $v_2$ for $k+1$.
\end{enumerate}

\nextcase $T \evalsto \forallT B$
\begin{gather*}
  \ru{B[a] = B[a] \in \SET[\ell] \mforall a \in \SIZE
    }{T \in \SET[\ell]}
\\[2ex]
  \ru{S_1 \evalsto \forallT{B_1} \qquad
      B[a] \shape B_1[a'] \mforall a, a' \in \SIZE
    }{T \shape S_1}
\\[2ex]
  \ru{S_2 \evalsto \forallT{B_2} \qquad
      B[a] \shape B_2[a'] \mforall a, a' \in \SIZE
    }{T \shape S_2}
\end{gather*}
\begin{enumerate}
\item To show $\up{S_1}{n_1} = \up{S_2}{n_2} \in T$ assume arbitrary $a_1,a_2 \in \SIZE$.
  Since $n_1 \ann{a_1} = n_2 \ann{a_2} \in \NE$, we obtain
  $\up{B_1[a_1]}{(n_1 \ann{a_1})} = \up{B_2[a_2]}{(n_2 \ann{a_2})} \in B[a_1]$
  by induction hypothesis.
  Thus, $(\up{S_1}{n_1}) \ann{a_1} = (\up{S_2}{n_2}) \ann{a_2} \in B[a_1]$
  by weak head expansion, which entails the goal by definition of $\EL(T)$.
\item Assume $k \in \NN$ and note that $\lev k = \lev k \in \SIZE$, hence,
  $t_1\,\ann{\lev k} = t_2\,\ann{\lev k} \in B[\lev k]$.  Thus, by induction hypothesis,
  $\R{k+1}{\down{B_i[\lev k]}(t_i\,\ann{\lev k})}{v_i}$, and finally
  $\R k {\down{S_i} t_i} {\lambda v_i}$ by definition of read back.

\end{enumerate}

\end{caselist}
\end{proof}

\begin{corollary}
  If\/ $T \in \SET[\ell]$ and $t = t' \in T$ then $\down T t = \down T {t'} \in \NF$.
\end{corollary}

\subsection{Fundamental Theorem}
\label{sec:fund}

PER of substitutions
\fbox{$\eta = \eta' \eeq \rho \in \Gamma$}.
\begin{gather*}
  \ru{
    }{\sempty = \sempty \eeq \sempty \in \cempty}
\qquad
  \ru{\eta = \eta' \eeq \rho \in \Gamma \qquad
      T\rho \in \SET[\ell] \qquad
      u = u' = t \in T \rho
    }{\sext \eta u = \sext{\eta'}{u'} \eeq \sext{\rho}{t} \in \cext \Gamma T}
\\[2ex]
  \ru{\eta = \eta' \eeq \rho \in \Gamma \qquad a \in \SIZE
    }{\sext \eta a = \sext{\eta'}{a} \eeq \sext{\rho}{a} \in \cext \Gamma \Size}
\qquad
  \ru{\eta = \eta' \eeq \rho \in \Gamma \qquad a,a',b \in \SIZE
    }{\sext \eta a = \sext{\eta'}{a'} \eeq \sext{\rho}{b} \in \erext \Gamma \Size}
\end{gather*}
We write $\rho \in \Gamma$ for $\rho = \rho \eeq \rho \in \Gamma$.
\begin{lemma}[Resurrection]
  \label{lem:resenv}
  If $\eta = \eta' \eeq \rho \in \Gamma$ then $\rho \in \resurrect\Gamma$.
\end{lemma}

Semantic judgements:
\[
\begin{array}{l@{\quad}l@{\quad}l}
  \models () & \defas & \mtrue \\
  \models \eext \Gamma \Size & \defas & \models \Gamma \\
  \models \cext \Gamma T     & \defas & \models \Gamma \mand \resurrect\Gamma \models T
\\[2ex]
  \Gamma \models s & \defas & \models \Gamma \\
%  \Gamma \models s & \defas & \mtrue \\
  \Gamma \models T & \defas & \Gamma \models T = T \\
  \Gamma \models T = T' & \defas & \Gamma \models T = T' : s \mforsome s %\Set[\ell] \mforsome \ell
\\[2ex]
  \Gamma \models t : T & \defas & \Gamma \models t = t : T \\
  \Gamma \models t = t' : T & \defas & %\models \Gamma \mand
    % \resurrect\Gamma \models T \mand
    \models \cext \Gamma T \mand
    t \eta = t' \eta' \in T \rho \mforall \eta = \eta' \eeq \rho \in \Gamma
\\[2ex]
  \Gamma \models \sigma : \Delta & \defas & \Gamma \models \sigma = \sigma \eeq \sigma : \Delta \\
  \Gamma \models \sigma = \sigma' \eeq \tau : \Delta & \defas &
    \models \Gamma \mand \models \Delta \mand
    \sigma \eta = \sigma' \eta' \eeq \tau \rho \in \Delta
    \mforall \eta = \eta' \eeq \rho \in \Gamma
\end{array}
\]

\begin{theorem}[Fundamental theorem]
  \bla
  \begin{enumerate}
  \item
  If $\der \Gamma$ then $\models \Gamma$.
  \item
  If $\Gamma \der J$ then % $\models \Gamma$ and
  $\Gamma \models J$
  \end{enumerate}
\end{theorem}
\begin{proof}
Simultaneously, by induction on the derivation.

\begin{caselist}
\nextcase $\forall$-introduction.
\[
  \ru{\erext \Gamma \Size \der t = t' : T
    }{\Gamma \der \lambda t = \lambda t' : \forall T}
\]
First $\models \Gamma$ follows from the induction hypothesis $\models {\erext \Gamma \Size}$.
To show $\resurrect\Gamma \models \forall T$,
assume $\eta = \eta' \eeq \rho \in \resurrect \Gamma$
and show $(\forall T)\eta = (\forall T)\eta' \in \SET[\ell]$ for some $\ell$.
To this end, assume $a \in \SIZE$ and show $T(\eta,a) = T(\eta',a) \in \SET[\ell]$.
Note that $(\eta,a) = (\eta',a) \eeq (\rho,a) \in
\resurrect {\erext \Gamma \Size} = \cext {\resurrect\Gamma}\Size$,
thus, we can instantiate the induction hypothesis and obtain our goal.

For the main goal,
assume $\eta = \eta' \eeq \rho \in \Gamma$ and
show $(\lambda t)\eta = (\lambda t')\eta' \in (\forall T)\rho$.
To this end, assume arbitrary $a,a',b \in \SIZE$ and show
$t(\eta,a) = t'(\eta',a') \in T(\rho,b)$.
Since $(\eta,a) = (\eta',a') \eeq (\rho,b) \in \erext \Gamma \Size$,
we conclude by induction hypothesis.

\nextcase $\forall$-elimination.
\[
  \ru{\Gamma \der t = t' : \forall T \qquad
      \Gamma \der a, a' : \Size \qquad
      \resurrect\Gamma \der b : \Size
    }{\Gamma \der t \ann a = t' \ann {a'} : T[b]}
\]
First, $\models \Gamma$ follows by induction hypothesis.
For goal $\resurrect \Gamma \models T[b]$,
assume $\eta = \eta' \eeq \rho \in \resurrect\Gamma$ and show
$T[b]\eta = T[b]\eta' \in \SET[\ell]$ for some $\ell$.
By induction hypothesis,  $b\eta = b\eta' \in \SIZE$.
% , thus,
% $(\eta,b\eta) = (\eta',b\eta') \eeq (\rho,b\eta) \in \cext{\resurrect\Gamma}\Size$
By another induction hypothesis, $(\forall T)\eta = (\forall T)\eta' \in \SET[\ell]$,
which by definition entails our goal $T(\eta, b\eta) = T(\eta', b\eta') \in \SET[\ell]$.

For the remaining main goal,
assume $\eta = \eta' \eeq \rho \in \Gamma$ and
show $t \ann a \eta = t' \ann{a'} \eta' \in T[b]\rho$.
By definition of substitution, it suffices to show
$t \eta \ann{a \eta} = t' \eta' \ann{a' \eta'} \in T(\rho, b \rho)$.
By induction hypothesis,
$t \eta = t' \eta' : (\forall T) \rho$, thus, by definition of this PER,
$t \eta \ann {a_1} = t' \eta' \ann{a_2} \in T(\rho, b')$ for any size values $a_1$, $a_2$, and $b'$.
We conclude by choosing $a_1 = a \eta$ and $a_2 = a' \eta'$ and $b' = b \rho$.

We now argue that this choice is possible,
namely that $a \eta, a'\eta, b\rho \in \SIZE$.
Note that the induction hypothesis gives us
$a \eta  = a  \eta' \in \SIZE$ and
$a' \eta = a' \eta' \in \SIZE$.
By resurrection (Lemma~\ref{lem:resenv}) we have $\rho \in \resurrect\Gamma$,
thus, by induction hypothesis, $b \rho \in \SIZE$.
\end{caselist}
\end{proof}

\section{Completeness of NbE}
\label{sec:compl}

% \begin{lemma}[Typing of constructed data]
% \label{lem:invdata} \bla
% If\/ $\Gamma \der c : \Nat b$ then $\resurrect\Gamma \der b : \Size$ and $b > 0$.
% \end{lemma}
